---
title: "CATaN: Canonical correlation Analysis of Transcriptome and TF-gene regulatory Networks"
author:
  - name: Haruka Takahashi, Hiroaki Hatano, Kazuyoshi Ishigaki
package: CATaN
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{CATaN Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = TRUE,
    warning = FALSE,
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

**CATaN** (Canonical correlation Analysis of Transcriptome and TF-gene regulatory Networks) identifies shared regulatory axes between transcription factor (TF)
binding patterns and transcriptome profiles. The resulting Canonical Component
(CC) scores are mapped to genomic SNP positions, producing annotation files
suitable for stratified LD Score Regression (S-LDSC).

## Workflow overview

1. **Filter and normalise** RNA-seq count data
2. **Run CCA** between a TF-gene regulatory Networks (TF-GRN) matrix and the transcriptome
3. **Align CC scores to SNPs** using TF ChIP-seq peak coordinates
4. **make binary annotations** and export for S-LDSC

# Quick start

```{r load-library}
library(CATaN)
```

## Step 1: Prepare input data

CATaN requires two matrices with shared gene identifiers as row names:

- **TF-GRN matrix**: a prepared matrix (gene × TF) indicating the regulatory relationships between TFs and their target genes made based on TF ChIP-seqs
- **RNA-seq count matrix**: raw counts (gene × sample).

```{r download-data, eval=FALSE}
## Download TF-gene binding matrix from Zenodo
tf_matrix <- download_tf_matrix("~/catan_data")
tf_matrix <- as.matrix(tf_matrix)
```

For this tutorial we use simulated RNA-seq data:

```{r simulate-data}
set.seed(42)
ngenes <- 500
ntf    <- 30
nsamp  <- 50

## Simulated TF-gene binding matrix
tf_matrix <- matrix(sample(0:1, ngenes * ntf, replace = TRUE,
                           prob = c(0.7, 0.3)),
                    nrow = ngenes, ncol = ntf)
rownames(tf_matrix) <- paste0("ENSG", sprintf("%05d", seq_len(ngenes)))
colnames(tf_matrix) <- paste0("TF", seq_len(ntf))

## Simulated RNA-seq counts
counts <- matrix(rnbinom(ngenes * nsamp, mu = 50, size = 2),
                 nrow = ngenes, ncol = nsamp)
rownames(counts) <- paste0("ENSG", sprintf("%05d", seq_len(ngenes)))
colnames(counts) <- paste0("sample", seq_len(nsamp))
```

## Step 2: Filter low-expression genes

```{r filter}
filtered_counts <- filter_low_expression(counts)
```

## Step 3: Prepare matrices

Find common genes between the count data and TF matrix, and perform TMM
normalisation:

```{r prepare}
mats <- prepare_matrices(filtered_counts, tf_matrix)
names(mats)
dim(mats$tf)
dim(mats$expr)
```

## Step 4: Run CCA

```{r run-cca}
cca_result <- run_cca(
    tf_matrix     = mats$tf,
    transcriptome = mats$expr,
    n_cc          = 10L,
    n_hvg         = 10000L
)
cca_result
```

## Step 5: Inspect results

### CCA parameters

```{r parameters}
ccParameters(cca_result)
```

### tfsampleLoading

```{r tfsampleLoading}
head(tfsampleLoading(cca_result))
```

### Visualisation

```{r plot-cc, fig.width=6, fig.height=5}
if (requireNamespace("ggplot2", quietly = TRUE)) {
    plotCC(cca_result, cc_x = 1, cc_y = 2, space = "tf")
}
```

# SNP annotation for S-LDSC

After CCA, CC scores are mapped to genomic SNP positions using TF ChIP-seq
peak BED files. This section describes the workflow; it requires external
data files not included in this vignette.

## Download reference data

```{r download-peaks, eval=FALSE}
## Download TF peak BED files from Zenodo (first time only)
peak_dir <- download_peak_beds("~/catan_data/peaks")
```

## Prepare SNP coordinates

Target SNPs should be provided as a `GRanges` object. For users with
1000 Genomes Phase 3 `.bim` files, a convenience function is provided:

```{r bim-example, eval=FALSE}
## Convert .bim file to GRanges
snp_gr <- bim_to_granges("path/to/1000G.EUR.QC.22.bim")
snp_gr

## Or combine all chromosomes
bim_files <- list.files("path/to/bim_dir", pattern = "\\.bim$",
                        full.names = TRUE)
snp_list <- lapply(bim_files, bim_to_granges)
snp_gr <- do.call(c, snp_list)
```

## Align CC scores to SNPs

```{r align-example, eval=FALSE}
library(BiocParallel)

annot <- align_cc_to_snps(
    catan_result = cca_result,
    peak_dir     = peak_dir,
    snp_gr       = snp_gr,
    chromosomes  = paste0("chr", 1:22),
    BPPARAM      = MulticoreParam(4)  # 4 cores
)
```

## make CC annotations and export

```{r export-example, eval=FALSE}
## Score 1 for top and bottom 10% of SNPs by CC score, and score 0 for other SNPs
annot <- extract_top_bottom_snps(annot, percentile = 0.1)

## Export as sLDSC-ready BED files
export_for_sldsc(annot, output_dir = "sldsc_annotations/", prefix = "CATaN")
```

Output files:

```
sldsc_annotations/
├── CATaN_CC1_rotation_top.bed.gz
├── CATaN_CC1_rotation_bottom.bed.gz
├── CATaN_CC2_rotation_top.bed.gz
├── CATaN_CC2_rotation_bottom.bed.gz
└── ...
```

These files can be used directly as annotations in `ldsc.py`.

# End-to-end wrapper

For convenience, `run_catan()` executes the entire pipeline:

```{r wrapper-example, eval=FALSE}
result <- run_catan(
    counts      = counts,
    tf_matrix   = tf_matrix,
    peak_dir    = peak_dir,
    snp_gr      = snp_gr,
    n_cc        = 10L,
    n_hvg       = 10000L,
    percentile  = 0.1,
    output_dir  = "sldsc_output/",
    BPPARAM     = MulticoreParam(4)
)
```

# Session information

```{r session-info}
sessionInfo()
```
